---
title: 'Lab 3: Data Collection & Introduction to Inference '
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---

## Before you start:

1.  Complete the Lab Primer linked in Canvas.  
2.  Be sure you have opened the `lab3-intro-ci-student.Rproj` file.  
3.  Change the name of the **QMD** file only to your last name, first initial, e.g., `OtterM-lab3-intro-ci-student-f26.qmd`. You will have to close the QMD document to rename the file.
4.  Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter"



## Skills

-   Describe and identify sampling techniques and any potential biases from them.  
-   Write a statistical research question which clearly defines the purpose, population, parameter, and variable of interest.  
-   Describe the sampling distribution of a statistic and define the standard error of a statistic. Be able to distinguish the standard deviation from the standard error.  
-   Calculate the standard error.  
-   Given a study, describe the sampling distribution of $\bar{x}$ as specifically as possible.  
-   Explain that in order to estimate the population parameter, we must quantify the uncertainty from our point estimate.  
-   Identify and define the population and parameter of interest in a given study.  
-   Construct and interpret a confidence interval in the context of the problem.  
-   Draw conclusions that apply to the population based on the results of the confidence interval; interpret the information in context.  
-   Make graphs in R to represent a numeric variable.

These skills correspond to LTs 1, 2, 5, and 6, and R LT R2.
*Questions will be labeled with their corresponding LT. Note that any one question may not fully cover the LT as it will be assessed on exams. Be sure to refer to the examples in your note outlines. Other questions may not correspond to a LT, but are foundational for learning them.* 



***


**Reminder**: Before starting this lab, let us load the `tidyverse`, `ggformula` and `mosaic` packages we need for statistical and graphical functions. Use the `library()` function to load in each package in the code chunk below.

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(ggformula)
library(mosaic)
```


***


## Part 1: Data Collection: Southern bull-kelp

A researcher was interested in examining long-distance dispersal of drifting southern bull-kelp (*Durvillaea antarctica*) adults in the New Zealand Canterbury Bight to understand connectivity between coastal populations. (Collins et al. 2010. Mol. Ecol.) (Invertebrates can 'raft' on the detached kelp, which also remains alive until beach-cast, allowing long-distance dispersal events to occur.) Given part of their written methods below, answer the questions that follow.

**Methods** (excerpt) Collection of beach-cast *D. antarctica* samples from seven field sites spanning the Canterbury Bight region was performed in March 2009 and again in April/May 2010. At individual field sites, we typically sampled across approximately 0.5-2.0 km of coastline. Where possible, we sampled only the beach-cast specimens that included a holdfast and/or primary stipe to ensure that individuals were not sampled more than once. In the cases (primarily in 2010) where relatively low numbers of intact specimens were found at some sites, we collected tissue from large pieces of *D. antarctica* that were clearly separated (e.g. >20 m) from other such pieces, again to avoid repeated sampling of individual specimens. We generally targeted specimens stranded at or above the high tide mark and sampled the dried tips of their fronds to maximize success with DNA extraction. Typically, between 20 and 30 samples were collected per site per year, with a small piece (a few cm) of frond tissue preserved in 95% ethanol before being transferred to silica gel.

For all multiple choice questions, **bold** (or **bold**) your answers but do not include the beginning letter or ending double spaces.


### 1.1.

(LT 1) What is the population of interest in this study?

a.  all drifting *D. antarctica* in the Canterbury Bight  
b.  all beach-cast *D. antarctica* in New Zealand
c.  all beach-cast *D. antarctica* in the Canterbury Bight  
d.  all *D. antarctica* in the Canterbury Bight


### 1.2.

What is one case in this study?

a.  one *D. antarctica*
b.  one month
c.  one coastline  
d.  one field site


### 1.3.

(LT 1) What type of sample design did the researchers use?

a.  Stratified Random Sample  
b.  Simple Random Sample  
c.  Convenience Sample  
d.  Volunteer Sample


### 1.4.

(LT 5) Biological studies can't always use a simple random sample like other fields, as they often don't have a sampling frame. How could the researchers introduce randomness into their study?

> Insert answer here


### 1.5.

(LT 5) If the researchers assumed that the kelp on the beach was already a random sample of the kelp drifting in the ocean, could the researcher generalize their sample to the population? Why or why not?

> Insert answer here



***

## Part 2: Sampling Distribution Properties

### 2.1.

Define the following terms:

> Population distribution:   
 
> Sample distribution:   

> Sampling distribution:   


### 2.2.

(LT 9) Describe the characteristics for a *general* sampling distribution of the sample mean from a population with mean $\mu$ and standard deviation $\sigma$. Be as specific as you can.

> Center:   

> Spread:   

> Shape:   


### 2.3.

Why is there variation around the center of the sampling distribution?

> Insert answer here


### 2.4.

(LT 9) Does the underlying population (the population from which you are taking samples) distribution affect the center of the sampling distribution? Explain why or why not.

> Insert answer here


### 2.5.

(LT 9) What factors influence the standard error for a sample mean? List each factor and how it influences the standard error if it increases.

> Factor 1:  
> Factor 2:



***

## Part 3: Data Collection - Song Length

We are interested in estimating the actual value of the length of recorded songs in your favorite musical genre. 


### 3.1.

Collect data from your own sample of *at least* 32 songs from your favorite music genre for the variable of song length (in seconds -- be sure to convert to all the same unit!) Decide how to take your sample using any sampling method we've learned and collect the data.

You are going to create your dataframe in R, rather than reading in a csv (better practice would be a csv, but we'll do this to make submissions/grading easier). Fill in the code below by replacing "_value_" with one case's song length in seconds. Add more values separated by commas. 

This will create a variable called `seconds` and store it inside the dataframe `songs`, which can be treated just like data you read in from a csv. 


```{r}
#| label: songs-data
songs <- data.frame(seconds = c(_value_, _value_, _value_, <add more as needed>)) # add your data here, writing your data values for each case separated by commas. You do not need to understand this code. 
head(songs)
```



### 3.2.

(LT 2, 5) What sampling method did you use? Describe and identify the sampling method. Any possible biases present? Justify your answers and be as specific as needed.

> Description of collection method:   
> Sampling Design Name & any biases:   
> Justification:  


### 3.3.

(LT R2) Create a graphic of your sample distribution, making sure to label it appropriately.

**Add code chunk here**


### 3.4.

(LT 3) Describe the shape of your distribution.

> Insert answer here


### 3.5.

(LT R1) Calculate the descriptive statistics that are most appropriate to quantify your distribution's center and spread. Write the values below using symbolic notation (e.g.: symbol = value, using Quarto to correctly format the symbols).

**Add code chunk here**

> Insert answer here




***

## Part 4: Inference: Estimating Song Length

We are interested in estimating the actual value of the length of recorded songs in your favorite musical genre. 


### 4.1.

What is the statistical research question?  

> Insert answer here


### 4.2.

(LT 1) What is the population of interest in this study? Remember to be as specific as needed.   

> Insert answer here


### 4.3.

(LT 6) What is the parameter of interest in this study?  

> [insert appropriate markdown symbol here] = [state parameter of interest in words]


### 4.4.

Fill in the table describing what each symbol represents, whether it represents a parameter or a statistic, and the calculated values of each in this context, if known. You may need to insert a code chunk to calculate the values that can be known. Type "NA" in the cell, if unknown.

| Symbol | Description | parameter or statistic? | Calculated Value if possible (with units) |
|--------|-------------|-------------------------|-------------------------------------------|
| $\mu$  |             |                         |                                           |
| $\bar{x}$ |          |                         |                                           |
| $s$ |                |                         |                                           |
| $n$ |                |                         |                                           |


### 4.5.

(LT 10) Check the conditions for constructing a confidence interval on your data. The conditions are listed, you just need to answer if they were met and justify why in the context of the question, including context or evidence from your study. You can replace the [hints].  

> 1.  Random sample: [Met?]; [Justification]  

> 2.  Normality or Sufficient sample size: [Met?]  
>     2a. Is the sample from a Normally distributed population or does the sample data have a roughly bell-shaped or symmetric distribution? [Yes / No]; [Justification]   
>     2b. Is the sample size sufficient? [Yes / No]; [Justification]  


### 4.6.

Let's use a model to find a 95% confidence interval.  
- Go to: http://www.rossmanchance.com/applets/2021/tbia/TBIA.html  
- Change the scenario to 'one mean'  
- Enter your sample's $n$, $\bar{x}$, and $s$ into the fields, then 'calculate'  
- Click the box next to "Confidence Interval".  
- Enter the correct confidence level.  
- Click 'Calculate.'  
- The confidence interval and degrees of freedom are displayed.

> Record the calculated confidence interval and degrees of freedom that is returned CI = ( ____ , ____ ), df = ____


### 4.7.

(LT 7) Interpret your confidence interval from 4.6 in the context of the question.

> Based on our sample, we are **[insert confidence level]** that the **[write parameter of interest]** is a single value between **[insert lower bound value]** and **[insert upper bound value]** **[insert units]**.


### 4.8.

(LT 4) Is there evidence from Exploratory Data Analysis that the length of recorded songs in your favorite genre more than 300 seconds? Be sure to cite the evidence, as well as describe why it supplies evidence for the value of the parameter.

> Insert answer here


### 4.9.

(LT 5) Based on the sample you took, can we generalize to the population of interest? Why or why not? If so, explain why. If not, explain why and, if applicable, the population that could be generalized to.

> Insert answer here


### 4.10.

(LT 11) To use the model to calculate the confidence interval, a condition/conditions from 4.5 needed to be satisfied. Which condition(s) are they, and what if you hadn't met that condition? What would that mean for your conclusion?

> Insert answer here


Before you finish -- check your knitted document. Is it easy to read? Do you need to add spaces anywhere to create line breaks? Did you accidentally delete any of the answer formatting?
