---
title: 'Lab 5: Two Group Comparison EDA & QQ Plots'
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---


## Before you start:

1. Complete the Lab Primer linked in Canvas.  
2. Be sure you have opened the `lab5-eda-qq-student.Rproj` file.  
3. Change the name of the **QMD** file only to your last name, first initial, e.g., `OtterM-lab5-eda-qq-student.qmd`. You will have to close the QMD document to rename the file. 
4. Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter"




## Skills

* Identify the explanatory and response variables in an experiment or observational study  
* Identify the type of study and its explanatory and response variables  
* Identify types of randomization in a study  
* Be able to produce appropriate graphics and summary statistics for independent two-sample and paired data  
* Be able to interpret data from a boxplot and histogram  
* Understand how a QQ plot is created and why it can assess normality
* Be able to assess normality of a sample based on the QQ plot  

The skills in this lab include LTs 3, 4, 13-17, and 20, in addition to R1 and R2.
*Questions will be labeled with their corresponding LT. Note that any one question may not fully cover the LT as it will be assessed on exams. Be sure to refer to the examples in your note outlines. Other questions may not correspond to a LT, but are foundational for learning them.* 



Here are some symbols that you might find useful:  

$\sigma$ 

$\mu$ 

$\mu_{word}$ 

$\mu_{d~word}$


***

Before you start the lab, add the code that will load any other packages you might need into this code chunk. 

```{r}
#| label: setup
#package for data management
#package for graphics
#package for stats

```



## Scenario 1: Fat Bears

Katmai National Park is famous for the salmon run each summer and the brown bear feast that accompanies it. Brown bears can lose roughly 30% of their body weight while hibernating, so it is essential that they pack on the pounds during the summer. They feed on high-caloric foods, like sockeye salmon (one salmon can be ~4000 calories), but also on sedges (a grass, often eating about 30 lbs of sedges a day), berries, grubs, etc. Over the course of the summer, bears have to add hundreds of pounds to their body weight to make up for weight lost during the past hibernation, and to survive the next hibernation; at peak season, brown bears can add up to four pounds of weight per day! Over the summer, the brown bears can really pack it on! (case in point: see pictures of the [Katmai Fat Bear champion  contenders](https://explore.org/meet-the-bears) from this fall's competition)

You are a researcher interested in the weight gain of brown bears over the summer. Each bear in the park is tagged with a unique identification number. You weigh a random selection of brown bears in Katmai NP at the beginning of June, and then weigh the same bears at the end of September and record their weight (in pounds). 

This code reads in our data and then prints a list of the vector names to the screen.

```{r}
#| label: load-bears
bears <- read_csv("fat-bear-weights.csv", show_col_types = FALSE)
names(bears)
```


### 1.1. 

(LT 13) Identify each variable of interest in our analysis and the type of variable for each by replacing the [placeholders] below.

Variable        | Explanatory or Response Variable?  | Numeric or Categorical?
----------------|------------------------------------|-------------------------
[placeholder]   | [placeholder]                      | [placeholder]           
[placeholder]   | [placeholder]                      | [placeholder]           



### 1.2. 

(LT 20) State the parameter(s) of interest in the study.

> Insert answer here  


### 1.3. 

(LT 14) State the study design of this study (two independent samples or one matched-pair sample). Justify your response.

> Insert answer here   
> Insert justification here  



### 1.4. 

(LT 13) State the study type of this study (observational or experiment?). Justify your response. 

> Insert answer here  
> Insert justification here  



### 1.5. 

(LT R1) Add code below to calculate the summary statistics (listed below) for *differences* in the weight gained over the summer. Place your resulting values in the table below, rounding to four decimal places as needed. 

```{r}
#| label: stats-bears
(gain_n <- length(bears$sept_lb)) #notice we just have to include one variable for n
(gain_xbar <- mean(~(____ - _____), data = ____)) #calculate mean difference
(gain_var <- _________________________) #calculate the variance of the difference
(gain_s <- ___________________________) #calculate the sd of the differences
```

Symbols     | Weight~diff~ 
------------|---------------
$n$         | [placeholder]              
$\bar{x}_d$ | [placeholder]              
$s_d^2$     | [placeholder]              
$s_d$       | [placeholder]   



### 1.6. 

(LT R2) Create a boxplot for the weight gained over the summer. Be sure to label the axes appropriately. 

< Insert code chunk here >  



### 1.7. 

(LT 3) Describe the distribution of weight gained over summer in brown bears based on your boxplot (e.g., center, spread, shape, outliers). 

< Insert code chunk here >  

> Insert answer here  



### 1.8 

(LT R2) Produce a QQ plot for the differences between September and June weights for the bears.

```{r}
#| label: qq-bears
gf_qq(~(_______ - ______), data = _________) |> 
  gf_qqline()
```



### 1.9 

(LT 16) Assess the plot for normality, including a justification.

> Insert answer here   



Note -- it would be *really* hard to actually weigh a grizzly bear; researchers at Katmai are working to develop new tools to measure bear volume with 3D laser imaging: [lidar in 2020](https://www.popsci.com/story/technology/fat-bear-week-laser-scanning-weight-estimates/) and [volume to weight ratios from 2019](https://medium.com/@exploreorg/how-big-are-fat-bear-week-bears-cd4f23036c33).



 
## Scenario 2: Yosemite Yellow-Legged Frog    

At one time the rare Sierra Nevada yellow-legged frog (*Rana sierrae*) was the most abundant amphibian in the high mountain lakes of the Sierra Nevada. Populations have since declined approximately 95% in the Sierra Nevada, including in Yosemite and on a trajectory toward extinction. Multiple factors have contributed to the decline of the frog, including the introduction of fish into naturally fishless waters throughout most of the park. Non-native trout feed on eggs, tadpoles, and young adults and compete for natural food sources; subsequently, no frogs live in lakes with fish. Although fish stocking in Yosemite ceased in 1991, many non-native trout populations continue to exist. The ecological effects of the loss of the Sierra Nevada yellow-legged frog from most of its range have been significant, as their former abundance made them a keystone predator and prey and a crucial agent of nutrient and energy cycling in Sierra Nevada aquatic and terrestrial ecosystems. 

Yellow-legged frogs still exist in habitats that have remained naturally fishless. Scientists in Yosemite have been removing non-native fish from some remote lakes and streams and monitoring the frog population to see if it can recover in these restored fishless habitats. They are interested in determining if the density of yellow-legged frogs in restored habitats five years after the restoration is still less than the density in natural habitats. From a map of all frog habitats in Yosemite, they randomly select locations along the shallows and near-edges in restored and in natural habitats. In each location, they calculated the density (#/m^2^) of adult frogs captured and record the habitat status. For the order of the difference, use `restored - natural`.    


### 2.1. 

Insert a code chunk that will read in your csv file `yellow-legged-frogs.csv`, save it as the object `frog`, and then print a list of the vector names to the screen. (*Note: You can __NOT__ use `View`*.)

< Insert code chunk here >  



### 2.2. 

(LT 13) Identify each variable of interest in our analysis and the type of variable for each by replacing the [placeholders] below.

Variable           | Explanatory or Response Variable?  | Numeric or Categorical?
-------------------|------------------------------------|-------------------------
[placeholder]      | [placeholder]                      | [placeholder]           
[placeholder]      | [placeholder]                      | [placeholder]           



### 2.3. 

(LT 17) State the parameter(s) of interest in the study.

> Insert answer here  


### 2.4. 

(LT 14) State the study design of this study (two independent samples or one matched-pair sample?). Justify your response.

> Insert answer here   
> Insert justification here  



### 2.5. 

(LT 13) State the study type of this study (observational or experiment?). Justify your response. 

> Insert answer here    
> Insert justification here  



### 2.6. 

(LT R1) Calculate the relevant summary statistics. Annotate your code using `#` to describe what each line of code is doing. Enter your values in the table below.

< Insert code chunk here >  


Symbols   |   Natural     |  Restored        
----------|---------------|---------------
$n$       | [placeholder] | [placeholder]              
$\bar{x}$ | [placeholder] | [placeholder] 
$s^2$     | [placeholder] | [placeholder] 
$s$       | [placeholder] | [placeholder] 



### 2.7. 

(LT R2) Create a boxplot of each group in your sample in one single graphic (i.e., numerical variable by categorical variable). Be sure to label the axes appropriately.

< Insert code chunk here >  



### 2.8. 

(LT 4) Without conducting the hypothesis test, what might you conclude about the about the density of yellow-legged frogs (measured as  $number / m^2$)) for natural habitats and for restored habitats, based solely on the provided exploratory data analysis in the previous questions? Justify your conclusion about the parameter(s) using the graphics and summary statistics. 

> Insert answer here      


### 2.9. 

(LT R2) Produce a QQ plot of the density of yellow-legged frogs for each habitat type (natural / restored). 

< Insert code chunk here >   



### 2.10. 

(LT 16) Assess each plot for normality, including a justification. 

> Insert answer for first graphic here     

> Insert answer for second graphic here  




