---
title: 'Lab 4: Confidence Intervals for One Population Mean'
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---


## Before you start:

1. Complete the Lab Primer linked in Canvas.  
2. Be sure you have opened the `lab4-confidence-student.Rproj` file.  
3. Change the name of the **QMD** file only to your last name, first initial, e.g., `OtterM-lab4-ci-student.qmd`. You will have to close the QMD document to rename the file. 
4. Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter"


## Skills

*  To recognize and pull out key information from a research scenario  
*  Construct a confidence interval for a population mean  
*  Given a confidence level ($1-\alpha$), determine the critical value (t*) needed to construct the confidence interval   
*  Construct and interpret a one sample interval for the mean based on the t-distribution  
*  Assess the conditions necessary to construct a Student's t confidence interval

These skills correspond to LTs 1, 2, 6, 7, 8, 10, 11, and 12, and R LT R1-R3.
*Questions will be labeled with their corresponding LT. Note that any one question may not fully cover the LT as it will be assessed on exams. Be sure to refer to the examples in your note outlines. Other questions may not correspond to a LT, but are foundational for learning them.* 


***

**Reminder**: Before starting this lab, let us load the `tidyverse`, `ggformula` and `mosaic` packages we need for statistical and graphical functions. Use the `library()` function to load in each package in the code chunk below.

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(___________) # add in the package with our graphics functions
library(___________) # add in the package with our statistics functions
```


***

## Part 1: Run differential  

Baseball is a game for statistics lovers. The run differential is defined as the number of runs between the winning and losing teams' scores. For example, if the Giants score 5 and the Dodgers score 7, the run differential of this game is -2 from the Giants’ perspective. One particular statistic is the mean run differential. You'd like to know if your favorite team is, on average, winning games (having a mean run differential greater than zero). Suppose you recorded the run differential for your favorite team for a random sample of games throughout the season. Your data are in a vector called `Run.Diff` in the file `run-differential.csv`. Load in the data below.  

```{r}
#| label: rundiff-data

baseball <- _____(______)
```


### 1.1.

(LT R1, R2) Before conducting any inference, such as a confidence interval, it is recommended that we perform exploratory data analysis. In the code chunk below, calculate the sample size, mean, and standard deviation and create a boxplot of the data in `baseball` for `Run.Diff`. For an appropriate graphic, do not forget to label axes for readers. Then also write the statistics using mathematical notation, and describe them verbally in context, below the code chunk.  

< Insert code chunk here >  

> Insert answer here  



### 1.2.

(LT 10) Have the conditions been met to construct a Student's t confidence interval? The conditions are listed, you just need to answer if they were met and justify why in the context of the question, including context or evidence from your study, by replacing the [prompts] (including ]s).  

< Insert code chunk as needed here >  

>  1. Normality or Sufficient sample size:  [Met?]  
      1a. Is the sample from a Normally distributed population or does the sample data have a roughly bell-shaped or symmetric distribution? [Yes / No]; [Justification]       
      1b. Is the sample size sufficient? [Yes / No]; [Justification]  


 
### 1.3.

(LT R3) Calculate the 95% confidence interval, using the `t.test()` function. Modify the code below by replacing the blanks.  

```{r}
#|label: rundiff-CI

t.test(______, data = ______, conf.level = ______)$conf.int
```


 
### 1.4. 

(LT 7) Interpret your confidence interval in the context of the scenario.    

> Insert answer here  



### 1.5. 

(LT 8) Based on your confidence interval, is there evidence that your favorite team is winning on average (having a mean run differential greater than zero)?  

> Insert answer here  



### 1.6. 

(LT 5 & 9) Suppose you decided to take a second sample to revisit and revise your question. This time, instead of selecting *any* game throughout the season, you randomly select 10 games *only* from games played at your favorite team's stadium and calculate a confidence interval for the true mean run differential.  

Consider the two sampling methods: (i) You randomly sample 35 games throughout the season, (ii) You randomly selected 10 games played at your favorite team's stadium throughout the season. Discuss the consequences of these two sampling methods by answering: If possible, what population can you generalize to? How does the sample size affect your results?  

> Insert answer here  



***

## Part 2: Eutrophication & Dissolved Oxygen 

Scientists are concerned about eutrophication -- excess nutrients in water systems which cause large algal blooms and subsequent spikes in animal respiration. The cycle they are concerned with is this: (1) high nutrients in the water allow algae populations to increase in population sizes rapidly, then die off in large numbers at once. (2) Decomposers (organisms that decompose dead organic matter) multiply rapidly to feed on the dead algae. (3) Those unusually large populations of decomposers use up the oxygen molecules that are dissolved in the water through cellular respiration (i.e.: breathing). 

Scientists are especially concerned about estuaries, many of which locally are surrounded by agricultural land that create water runoff with excess nutrients from fertilizers. This high-nutrient water runs off the fields and into the estuary. Respiration spikes have the potential to lower dissolved oxygen (dO) in the water and potentially cause die-offs of animals if the dissolved oxygen gets too low. The EPA sets the minimum safe level for dissolved oxygen at 4 mg/L for estuarine fish species. Researchers are interested in determining the mean dissolved oxygen level in the Salinas River estuary. The researchers select 20 days in the year, after giving the days equal probability to be selected, to collect dO measurements in mg/L in the Salinas River estuary. The data they collected are in the variable `dO` (lower case d, upper case o, for oxygen (not 0 / zero)) inside the file `estuary-oxygen.csv.` Load in the data below. 

<Insert code chunk here>


### 2.1.

In the text above, bold the general research question. What is the *statistical* research question? Write it below & be specific.

> Insert answer here.  


### 2.2.

(LT R1, LT2) From the data, calculate the statistics listed in the following table, and then fill in the table with the statistics and parameters. If a parameter is unknown, write "NA." You can round to four decimal points as needed to enter the values. 

```{r}
#| label: dO-stats

# calculate statistics, annotating your code to describe what descriptive statistic each line calculates

```  
  
Symbol         | Value        
-------------- | ----------- 
$\mu$          |          
$\bar{x}$      |     
$\sigma$       |     
$s$            |     
$n$            |     
$df$           |     


### 2.3.

(LT R2) Create an appropriate graphic of the dissolved oxygen levels in the Salinas River estuary. Be sure to label your axes fully!

< Insert code chunk here > 

> Insert answer here.  



### 2.4. 

(LT 6) State the parameter of interest, verbally and symbolically.

> Insert answer here.   



### 2.5. 

(LT 10) Have the conditions been met to construct a Student's t confidence interval? State each condition, answer if it was met, and justify why in the context of the question.

< Insert code chunk as needed here >

> [Condition name]: [Decision]
    a. [evidence and justification in context of question]  
    b. [evidence and justification in context of question]


### 2.6. 

(LT R3) Calculate a 95% confidence interval for the mean dissolved oxygen level in the Salinas River estuary, then write your interval below. (Challenge yourself to write the answer in in-line code!)  

<Insert code chunk>

> Insert answer here.  



### 2.7. 

(LT 7) Interpret your confidence interval in the context of the problem. 

> Insert answer here.   



### 2.8. 

(LT 8) Adult fish that live in estuaries can withstand dO levels that are 4.0 mg/L or higher. Based on the confidence interval, is there evidence that Salinas River estuary dO levels are healthy for adult fish (average dissolved oxygen level greater than or equal to 4.0 mg/L)? Explain your reasoning in context.

> Insert answer here.  



### 2.9. 

(LT 8) Juvenile fish that live in estuaries can withstand daily dO levels that remain equal to or greater than 6.0 mg/L. Based on your confidence interval, is there evidence that the Salinas River estuary dO levels are safe for juvenile fish to survive, on average? Explain your reasoning in context.  

> Insert answer here.  



### 2.10. 

(LT 4) Provide a second piece of evidence from EDA (Exploratory Data Analysis) that supports the results of the confidence interval in answering whether the dissolved oxygen levels are safe for juvenile fish (i.e. average dO levels less than 6.0 mg/L).  

> Insert answer here.  


### 2.11. 

(LT 11) Can you trust the calculations of your confidence interval are accurate? Justify your answer with statistical reasoning.  

> Insert answer here.  


### 2.12. 

(LT 5) Identify the population, if any, to which the study can be generalized. Explain why with evidence.    

> Insert answer here.  


### 2.13. 

(LT 12) For each of the following, indicate if it is a correct or incorrect interpretation of the confidence interval you calculated in 2.6. Provide an explanation for each.

a. There is a 99% chance that the true mean dissolved oxygen level for days in the Salinas River estuary is between the upper and lower bounds of the confidence interval.

> Correct or Incorrect?  
> Explain.  


b. 99% of the sample dissolved oxygen levels for days in the Salinas River estuary will be between the upper and lower bounds of the confidence interval. 

> Correct or Incorrect?  
> Explain.  


c. Suppose we repeat this study, sampling and calculating a 99% confidence interval many times. 99% of the time, the true mean dissolved oxygen for days in the Salinas River estuary will be between the upper and lower bounds of the confidence interval. 1% of the time, it will not.

> Correct or incorrect?  
> Explain.  



### 2.14. 

(LT 9, 12) All else being equal, if the researchers repeat simple random sampling from the same population, how would the following changes influence the width of a confidence interval? (For each, choose one answer from the brackets and **bold** it, then write a *brief* justification below.    

a. Increased confidence level (1-$\alpha$): [Wider, Narrower, No Influence]  

> Justification  


b. Smaller sample size (n): [Wider, Narrower, No Influence] 

> Justification  


c. Larger sample standard deviation: [Wider, Narrower, No Influence]  

> Justification  






